Draft Article

### 2019-11-26 article.tex
Theory draft sent to Wohlforth

\subsection{Introducing the empirical examples}
% I use defensive realism to formulate a theoretical foundation to explain the existence of espionage norms generally, and the Soviet Union's extension of espionage norms to spy satellites in particular. Between two security-seeking states, both states will be less secure in the event of an arms race, so preventing one is a mutually-agreeable outcome. Resolving the security dilemma requires that the states signal their benign intentions, which states are more likely to do if they have good intelligence---intelligence that informs them about their adversary's capabilities and intent. Therefore, cooperating to preserve the practice of espionage as a semi-legitimate state institution makes rational sense, both because of the benefits espionage offers the state doing the spying, and the benefits the same state derives \emph{from being spied upon}.

>>>>

% States now recognize that some forms of espionage facilitate international cooperation.\footcite{baker_tolerance_2004}

A complete theoretical justification for espionage---that maintaining a mutual exchange of illicit information is better for both states than the alternative---can be constructed by combining the work of previous scholars, and finding archival evidence to support their claims. Christopher Baker explains how espionage makes functional cooperation possible through treaties, allowing states to verify each others' compliance.\footcite[p.~6]{baker_tolerance_2004} Charles Glaser uses the 1972 Anti-Ballistic missile treaty as an example of precisely the type of signal that could lead states to resolve the security dilemma, but he does not acknowledge that the intelligence operations of both states are what make signals like it possible.\footnote{It is entirely possible that Glaser himself realizes this, it's just not written down; the purpose of the book is to defend the theory, not explain the mechanics.} And Julius Stone, a professor of international law, foresaw this exact role for espionage in 1962, well before the public had any real idea about the US spy satellite program: \enquote{If you do not have a system of international inspection and if you can't get one (and I think it is quite likely that we can't) then the function which international inspection is supposed to serve still needs fulfilling. You may not be able to fulfill it to the optimum extent by reciprocally tolerated espionage, but you may be able to reduce the dangers.}\footcite[p.~41]{stone_legal_1962}

In 1962, Stone defended espionage from a practical perspective in exactly the same terms that I am now defending it from a theoretical one. Permitting at least some kinds of espionage is actually beneficial for both sides because it allows us to independently verify the intent and capabilities of other nations. Doing so actually reduces the likelihood that miscalculation will lead to open hostilities. As a result, both states allow it to happen, because they each benefit from a less-tense security environment. Stone aptly described this phenomenon as \enquote{reciprocally tolerated espionage.}

\subsection{Introducing the empirical examples}

The best way to prove that this theory has explanatory power for state decision-making is to find examples from the past where states made decisions that showed they were willing to accept the cost of enemy espionage in order to practice espionage themselves. In the ideal scenario, we would find evidence of policymakers justifying their decisions along these lines. Such decisions were made frequently during the Cold War, and with the explicit intent to increase the bilateral flow of information to create a more secure world environment. This article will present two such cases: aerial espionage and spy satellites.

Both planes and satellites played an underexplored role in the formulation of some of the Cold War's landmark diplomatic achievements. SALT I (the ABM Treaty) is widely considered to be the landmark diplomatic achievement of Nixon's \emph{d\'etente} policy, which set the stage for a broader easing of relations between the US and the USSR, and spy satellites were crucial to reducing the cost that each side would pay to sign it. Limiting ballistic missile defense systems was a relatively minor ask for both sides, but in the highly risk-averse context of the Cold War, the signal was significant.\footcite[p.~66]{glaser_rational_2010} Underneath the omnipresent threat of nuclear annihilation, \emph{d\'etente} provided a rational strategic choice for both states---made possible by their reciprocal tolerance of espionage.

% I have yet to find a quote from Nikita Khrushchev in which he states that it was good for the Soviet Union to be have its state secrets compromised. I don't expect that I ever will. The realist approach to espionage is difficult to acknowledge politically; it cuts against national pride and power in order to reach a pragmatic, somewhat emotionally unsatisfying conclusion. I asked a senior national security policymaker, Jake Sullivan, whether he thought American espionage was a net positive for global security. ``I wouldn't work for the American government if I did not believe that it was a net positive actor,'' he told me, ``so I want the United States to be able to make the most informed decisions possible.'' When I asked whether it promoted global stability for a foreign nation---Russia for instance---to spy on the United States, he qualified his answer.

% He split the types of intelligence that foreign state might gather on the United States into three types: intent, capabilities, and strategy. ``I have no problem with the Russians knowing that we do not have aggressive intent,'' he told me. Capabilities, however, are more complicated. Some defense capabilities would probably be fine to have another country independently verify, and some we might want to keep secret, such as the atomic bomb before the end of the Second World War. And under almost no circumstances would a state want to have its strategies and warfighting plans revealed, so those should be kept as secret as possible.\footcite{sullivan_personal_2019} Without invoking the theory itself, Sullivan described exactly the type of intelligence that a defensive realist would need to solve the security dilemma.\footnote{At the point in my research when I conducted this interview, I had never even heard the term ``defensive realism,'' lest you think I led him to it.}

% The problem is that a state will never be able to control what type of information is being leaked. Espionage is an all-or-nothing proposition; either a state is willing to risk compromising some government secrets---including the ones that it is always disadvantageous to leak---in order to preserve the right to attempt it itself, or the state defects and begins imposing severe diplomatic consequences for espionage, inevitably inviting the same consequences upon itself, at which point the information pipeline tightens drastically. Unfailingly, states will choose to continue cooperating, so they must all be making the same calculus that the information they lose is worth the security they gain from getting to practice espionage themselves.

% If you believe that American intent is benign, then tacitly permitting espionage is one way out of the security dilemma. Sullivan and Glaser did not conclusively explain, however, why \emph{every} major power continues to cooperate. More likely than not, some states today are not pure security-seekers. Glaser calls these greedy states, and when a state's motives are mixed between greed and security, those get assigned the label of greedy as well. For a greedy state, having another state correctly evaluate its intent could prove disastrous. The state might be pretending to have a benign intent in order to gain an advantage and espionage might reveal the truth.

% That is only partially true. I certainly believe that elements of espionage and its place in diplomacy are socially constructed. The practice of expelling diplomats, for instance, predates all the events discussed in this thesis, and almost certainly evolved from some sort of customary protocol, rather than a rational calculation about the most effective way of responding to espionage.

% An institutional theory of politics cannot explain the endurance of espionage norms, because there are no international institutions dedicated to governing espionage. As the legal section notes, peacetime espionage in most traditional sense---activities associated with human spies---has virtually no legal recognition at all. The first institutional recognition of espionage for conflict-reduction purposes is the SALT I treaty in 1972, which protected ``national means of verification,'' but espionage-as-verification is limited to the specific treaties in which it is in included. Espionage is probably just too uncouth to formally acknowledge.

% Likewise, the only legal regime that applied to a spy plane was the one that applied to planes in general, and that explicitly prohibited the U-2 overflights; the Treaty on Open Skies was signed in 1992, well after the events discussed here. The clear illegality of the U-2 overflights should have given the Soviet Union more leverage to gain concessions, not less. As for space, the Outer Space Treaty entered into effect in 1967, eight years after \emph{Discoverer 14}; the strange timing of that treaty, right as the USSR was developing ASAT capabilities, was noted in Chapter 4.

% I posit that the desire from all states to continue the practice of espionage writ large, even as they seek to stop particular instances of it against themselves, has an important psychological component. Every policymaker would like to their decisions to be as well-informed as possible. Without the benefit of hindsight, it is very difficult objectively gauge how much ``security'' is lost from a successful intelligence operation. By contrast, it is very easy for someone in charge of making an important decision to notice how much better off they are for having had access to some illicitly-obtained information. I don't mean to suggest that these trade-offs are always impossible to evaluate. Khrushchev was keenly aware of how the satellite photography would affect his missile bluff; preserving the missile bluff was just not worth impacting Khrushchevs's own ability to make use of satellite photography. In situations where it is difficult to evaluate relative gains, there exists a strong psychological bias towards preserving ones own view of the situation, even if it means allowing the adversary to have its view as well.

% If espionage often leads to mutually agreeable outcomes, then why is not just completely legal? Well, in some cases it is. A true Open Skies Treaty followed soon after the collapse of the Soviet Union. In just over a decade, spy satellites went from ``espionage in space'' to ``national technical means of verification.'' But even as espionage gained some legitimacy in statecraft, the vagueness of that wording, ``means of verification,'' is indicative of what keeps espionage from true public acknowledgment. The intelligence business is often looked at---not entirely unfairly---as being shady. Reframing espionage as treaty verification procedure can help mitigate that, but if formally acknowledging spy satellites feels uncouth, its hard to imagine that human spies will ever receive a legal regime of their own.

% Regardless of theoretical explanation, the cooperative outcomes of the Cold War is remarkable. The USSR had reason to believe that it would have benefited from a mutual prohibition on the technology that the United States was using. The next two chapters will go in-depth on the process that led to this decision---from both sides---to make the case that espionage norms have a fundamental appeal that the United States was able to exploit for its own benefit, and induce the Soviet Union to cooperate.

% I use defensive realism to formulate a theoretical foundation to explain the existence of espionage norms generally, and the Soviet Union's extension of espionage norms to spy satellites in particular. Between two security-seeking states, both states will be less secure in the event of an arms race, so preventing one is a mutually-agreeable outcome. Resolving the security dilemma requires that the states signal their benign intentions, which states are more likely to do if they have good intelligence---intelligence that informs them about their adversary's capabilities and intent. Therefore, cooperating to preserve the practice of espionage as a semi-legitimate state institution makes rational sense, both because of the benefits espionage offers the state doing the spying, and the benefits the same state derives \emph{from being spied upon}.

% \section{Aerial espionage}
% \section{Introduction}
% According to \emph{The New York Times}, the very first reconnaissance flight took place in 1911, only eight years after Kitty Hawk, when ``Lieut. Piazzi today, for the first time in the history of warfare, made from this place an aerial reconnaissance against a hostile power'' surveilling the Turkish Infantry near Tripoli.\footcite{special_cable_to_the_new_york_times_air_1911} The ability to fly over an enemy's position to gain intelligence about their movements---in a craft orders of magnitude more versatile than a balloon---was so obviously valuable that countermeasures quickly followed. Early aerial dogfighting famously took the form of pilots firing handguns at each other out of open cockpits in mid-air (emphasis on attempted), but by the time of the First World War, planes were already outfitted with synchronized machine guns and ground regiments equipped with anti-aircraft weaponry. Crucial information about military movements was going to come at a cost.

% Even though these aircraft were performing vital intelligence-gathering functions, that did not make them spy planes. Aerial reconnaissance first appeared in the press in 1911, but the term ``spy plane'' appeared in the pages of \emph{The New York Times} just once before 1960.\footnote{According to my search of the archives. The instance mentioned here was in 1944, when the AP wrote that ``Germans tried to probe the secrets of the gathering Allied storm by E-boat patrol dashes [...] and by spy-plane coastal raids.''(\cite{the_associated_press_britons_1944}) In context of the article, though, it's clear that they are referring to military aerial reconnaissance, especially since it was written during the Second World War. This is a marked difference from a plane created for peacetime intelligence, which is what the term ``spy plane'' has since come to mean.} The image of a spy plane crashed into the public consciousness that year when Gary Powers crawled out of a U-2 class aircraft over Soviet Russia. The downed craft was clearly not a generic fighter plane that had flown too close to the border, and it suddenly became necessary to categorize this new type of plane, flying at previously unheard-of altitudes, that was purpose-built for clandestine photography. The United States first claimed that the lost plane was a ``weather craft.''\footcite{caruthers_soviet_1960} When that cover didn't hold up, a new story was born. The diplomatic fallout of the U-2 shootdown, which threatened to torpedo an upcoming NATO-Warsaw summit, had become the ``spy-plane case.''\footcite[This is the second time that ``spy plane'' as a term of art appeared in \emph{The New York Times}. There would be many more.]{reston_allies_1960}

% I emphasize the linguistic distinction between a military plane used for reconnaissance and a ``spy plane'' because that distinction has a major impact on how the flight is perceived and what diplomatic repercussions might result when it is caught.\footnote{That is not to say there is a hard boundary between military and non-military intelligence, either in purpose or collection. The overlap between military and civilian intelligence agencies---especially in the United States---is considerable. The Air Force was deeply involved with the U-2 Project, for example, even though it was technically under the jurisdiction of the CIA. The aerial photographs taken by U-2 missions were used to inform both military and civilian operations.} A military fight carries with it the implications of combat, aggression, and the possible prelude to an attack. A spy flight is instead a tool of civilian intelligence, whose purposes are often frustrated but never quite discouraged. The creation of the U-2 plane---a craft exclusively for civilian intelligence---is a pivotal moment in the history of espionage because the U-2 is the physical embodiment of the espionage norms it helped to define. Its \emph{raison d'\^etre} is an activity which is not permitted---or its altitude would be unnecessary---but its design emphasizes its peaceful reconnaissance purpose.

% \section{The Open Skies policy}
% From the beginning of the Cold War, aerial intelligence played a crucial role in American foreign policy. The United States considered aerial surveillance so important that it was willing to allow the Soviets to do it in return. President Eisenhower formally proposed a mutual inspection regime called Open Skies to the Soviet Union as a means of reducing security tensions. In his prepared remarks at the Geneva Summit of 1955, Eisenhower laid out the following plan:

% 	% \begin{quote}
% 	\textbf{To give to each other a complete blueprint of our military establishments}, from beginning to end, from one end of our countries to the other, lay out the establishments and provide the blueprints to each other.
% 	\newline

% 	Next, to provide within our countries facilities for aerial photography to the other country---\textbf{we to provide you the facilities within our country, ample facilities for aerial reconnaissance}, where you can make all the pictures you choose and take them to your own country to study, you to provide exactly the same facilities for us and we to make these examinations, and by this step \textbf{to convince the world that we are providing as between ourselves against the possibility of great surprise attack}, thus lessening danger and relaxing tension. (emphasis mine)\footcite{eisenhower_president_1955}
% 	% \end{quote}

% The UN-sponsored Geneva Summit, formally the \emph{Conference on the Peaceful Uses of Atomic Energy}, was a comparatively hopeful period in the Cold War, a conference promoting scientific diplomacy where such a cooperative information exchange might have seemed feasible.\footcite[p.~27]{luscher_nuclear_2018} Nonetheless, the scale of this offer was astounding. Eisenhower offered to let the the Soviets install operational military bases in the United States from which they could surveil American national defense infrastructure, so long as they allowed the US to do the same.\footnote{Also included were reciprocal nuclear facility inspections, an immediate nonstarter in the USSR, but also controversial within the American administration back home. (\cite{prados_review_2015})} In effect, proposed a limited form of state-sanctioned espionage for explicitly defensive purposes---exactly the norm I am arguing exists in practice, albeit in a different form.

% The offer was quickly rejected. In conversation over buffet that evening, Khrushchev politely countered that the plan would not further disarmament at all because it would merely confirm the fragmentary information that their respective intelligence services already possessed.\footcite{department_of_state_memorandum_1955} The response was telling---though it was in no way an endorsement of the activity, it both acknowledged that espionage was taking place between the two superpowers, and that it served a potentially de-escalatory purpose.

% Khrushchev also claimed that the offer was primarily a propaganda move on the part of the US, but backed down when Eisenhower essentially dared him to call his bluff.\footcite[p.~2. This is a secondary source---the book cites a conversation by Eisenhower, quoted in NSC meeting memos that are stored in the Eisenhower library.]{bury_eisenhower_2014} The formal rejection came a few weeks later: ``While giving credit to this proposal's attempt to find a solution,'' USSR Council of Ministers chair Nikolai Bulganin wrote in his report to the Supreme Soviet, ``\textelp{} aerial photography cannot yield the expected results, since both our countries contain limitless expanses in which, if one desires, everything can be hidden.''\footcite{bulganin_meetings_1955} Though a lukewarm excuse, it was not a disdainful one. Earlier in the report, he had even said that the conference, as a whole, ``must be considered a definite success of the peace-loving forces.''

% Was the offer genuine, or was it the stunt that Khrushchev accused it of being? In a book about the relationship between aerial reconnaissance and Eisenhower's New Look foreign policy, historian Helen Bury argues that a willingness to negotiate acceptable and enforceable agreements with foreign powers was central to his Cold War strategy.\footcite[p.~69]{bury_eisenhower_2014} More importantly, the intelligence that an Open Skies policy promised would have been invaluable in his ongoing effort to control defense spending, where Eisenhower was embattled by deeply paranoid American institutions.\footcite[p.~212]{bury_eisenhower_2014} His commitment to Open Skies in practice, however, is still a point of some historical debate. In reviewing her argument, intelligence historian John Prados sees a contradiction between her portrayal of Eisenhower's genuine desire for arms control and Eisenhower's role in significant nuclear buildup.\footcite[p.~233-234]{prados_review_2015} In the following chapter, I will present my own evidence that these two existed as part of the same strategy. Eisenhower used photo-reconnaissance---of the kind that an Open Skies policy would have legally permitted---to create an accurate assessment the military capabilities required to sufficiently deter a Soviet nuclear strike; as a result, the United States built far fewer missiles than it otherwise would have.

% % Valentino: Actually, they were explicitly consistent. Ike thought that nukes were the cheapest way to counter the Soviet numerical advantage in conventional weapons (which were and are much more expensive). So Ike saw both as cost saving measures.

% Regardless of Eisenhower's intent, it is clear that the United States would have far more to gain from such a policy than the USSR, and it made strategic sense for the Soviets to reject the offer. Writing about the history of CIA's imaging program, former imagery analyst David Lindgren notes that Khrushchev had the advantage of huge amounts of publicly-available information about US military installations through newspapers, maps, and official aerial photographs, while Soviet restrictions on press freedom denied symmetrical data.\footcite[p.~38]{lindgren_trust_2000} Allowing for an Open Skies policy would also have revealed that Soviet air defenses were not nearly as far along as their leadership claimed. Khrushchev's son would later write that his father's overriding concern at the time was to conceal Soviet weakness.\footcite[p.~133]{brugioni_eyes_2010}

% Treaty or not, the United States was desperate for intelligence about the Soviet missile program, and aerial photo-reconnaissance was crucial to gaining it. The West's HUMINT capabilities within the Soviet Union were hamstrung by the intense surveillance under which their diplomats were placed, as well as the enormous size of the Russia itself.\footcite[p.~23]{lindgren_trust_2000} As the US increasingly relied on signals intelligence (SIGINT) to gather intelligence, the USSR began pursuing a more aggressive aerial defense policy, demonstrating their willingness to open fire on foreign reconnaissance aircraft. The increased cost associated with these missions did little to deter senior US policymakers from attempting to surveil the skies.\footcite[p.~4]{pedlow_central_1992}

% % The Soviet Union was able to successfully destroy US Military aircraft during purported peacetime because in many cases those US aircraft were not supposed to be there in the first place. In many cases their flight path was a clear violation of Soviet territorial boundaries---a fact of which the administration was aware.\footcite{goodpaster_memorandum_1956}

% In the rest of this chapter, I will argue that Eisenhower satisfied the need for intelligence that an Open Skies treaty would have provided by normalizing the collection of aerial intelligence and minimizing its diplomatic consequences. This process happened in two parts: first, through the routine resolution of military reconnaissance shootdowns, and second, by symbolically divorcing peacetime aerial espionage from military reconnaissance with the creation of a spy plane. For both the military and civilian missions, I will demonstrate how the even the most sensationalized incidents resulted in no more than ineffectual diplomatic protests, even though the spy plane could surveil locations that a ferret flight never could.

% \section{Military reconnaissance flights}
% \subsection{Timeline of early Cold War SIGINT missions}
% Well before the U-2 Incident in 1960, the United States had been routinely flying military aircraft near Soviet airspace, and the USSR had been routinely shooting them down. For a conflict defined by its lack of direct skirmishes between the two superpowers, a surprising number of Americans died in these incidents, which are mostly forgotten to history. A Smithsonian investigation found that over the course of the Cold War more than 200 American pilots were lost to Soviet shootdowns---126 of which remain unaccounted for to this day.\footcite{glenshaw_secret_2017}

% The purpose of these missions was to gather SIGINT identifying the location of critical radar installations along the Soviet border. Converted bombers---called ``ferrets'' by the Air Force---were outfitted with advanced radio equipment and sent to intercept as much information about Soviet radio transmissions as possible.\footcite[p.~4]{peterson_maybe_1993} According to the CIA, these missions began as early as 1947.\footcite[p.~4]{peterson_maybe_1993} Some of these flights only flew near the border, remaining over traditionally recognized international waters, while others were ``overflights,'' missions that deliberately violated Soviet airspace to collect intelligence at a much higher risk of failure. The earliest postwar overflight of the Soviet Union---of which I can find definitive record---took place on May 10, 1949, when two RF-80As surveilled the Kurile islands\footcite[p.~8]{peebles_shadow_2000}

% Determining the quantity of early Cold War reconnaissance flights is difficult because they were a closely guarded secret at the time. Some flights, however, were inadvertently declassified---when they were shot down by enemy air defenses. Since my goal is to understand the diplomatic repercussions of these flights, an effective place to start is a list of reconnaissance flights that attracted diplomatic attention after being shot down. The original shootdown stories are available in the newspaper archives, and the incidents have been helpfully compiled by US Air Force Academy historian Dr. John Farquhar, who specializes in aerial reconnaissance and wrote his master's thesis on how it shaped early Cold War diplomacy. My choice of incidents to analyze in this chapter is based in part on his work.
% % \newline

% Additionally, the CIA history magazine \emph{Cryptologic Quarterly} analyzed the Soviet shootdowns of US reconnaissance aircraft in depth, creating a comprehensive (as far as we know) list of each American overflight that was shot down by the USSR (Table \ref{soviet-shootdowns}). The article was written in 1993, but not declassified until 2009. The author, Michael Peterson, uses two limiting criteria that perfectly suit our purposes: the only flights listed are those that were (a) shot down by Soviet attacks, not Chinese, North Korean, etc., and (b) explicitly US reconnaissance aircraft.\footcite[p.~4]{peterson_maybe_1993} Note that not all of these flights are necessarily overflights; many took place over international waters when they were shot down. In effect, the table catalogs the 13 times in which the USSR shot down an American aircraft designed for collecting intelligence between 1950 and 1964---intercepted espionage attempts.\footcite[p.~5. In the original document, this table lists the first incident as having taken place over the Barents Sea, not the Baltic Sea. Because the description of the mission---including a map of its route in the same document---takes place entirely over the Baltic sea, I have concluded that this has to be a typographical error, and corrected it here.]{peterson_maybe_1993}

% \begin{table}[ht]
% \centering
% \begin{tabular}{llr}
% \textbf{Date}     & \textbf{\makecell[l]{U.S. Service \&\\ Aircraft Type}}   & \textbf{General Location} \\
% 8 April 1950      & USN PB4Y2 Privateer           & Baltic Sea                           \\
% 6 November 1951   & USN P2V Neptune               & Sea of Japan                         \\
% 13 June 1952      & USAF RB-29                    & Sea of Japan                         \\
% 7 October 1952    & USAF RB-29                    & East of Hokkaido/Kuril Is.           \\
% 29 July 1953      & USAF RB-50                    & Sea of Japan                         \\
% 4 September 1954  & USN P2V Neptune               & Sea of Japan                         \\
% 7 November 1954   & USAD RB-29                    & East of Hokkaido.Kuril Is.           \\
% 18 April 1955     & USAF RB-47                    & Off Kamchatka Peninsula              \\
% 10 September 1956 & USAF RB-50                    & Sea of Japan                         \\
% 2 September 1958  & USADF C-130                   & Soviet Armenia (near Turkish border) \\
% 1 May 1960        & CIA U-2                       & Sverdolsk, USSR                      \\
% 1 July 1960       & USAF RB-47                    & Barents Sea                          \\
% 10 March 1964     & USAF RB-66                    & East Germany
% \end{tabular}
% \caption{Summary of Soviet Shootdowns, 1950-1964}
% \label{soviet-shootdowns}
% \end{table}

% I cannot confirm whether Peterson's record is comprehensive. Regardless of how thorough declassified CIA information appears, I cannot rule out the possibility that there were other, more secret missions which for whatever reason the United States government has chosen to continue protecting to this day. Nonetheless, a few aspects of the \emph{Cryptologic Quarterly} article are promising, and suggest that the information is relatively complete. First, the article was written for a classified audience as an overview of reconnaissance shootdowns during the 1950s, so it is intended to be comprehensive. Shootdowns are a sufficiently significant event such that it is unlikely the author would have just missed one. Additionally, when the article was declassified 16 years later, it appears to have been declassified without redactions or missing pages. And as with any analysis of diplomatic repercussions and espionage, we are aided by a simple logical trap---if there had been another mission so highly classified that it was omitted from this document, it is difficult to imagine how diplomatic consequences could have been imposed with a degree of secrecy that would prevent us from knowing about it today.

% In the following section, I cross-reference the shootdowns that the CIA says took place with Farquhar's analysis of their media and diplomatic impact. A shootdown is either present in both lists, or just one of the two---some incidents in the CIA list are missing from Farquhar's list because they did not have a diplomatic impact, and some incidents that Farquhar analyzes are not mentioned by Peterson because they were not SIGINT reconnaissance missions and likely (though not definitively) served little espionage function. Together, the full spectrum of aerial reconnaissance incidents demonstrates a clear pattern in which the shootdowns that involved espionage were quickly resolved by both parties, while the shootdown of a flight unrelated to espionage produced swift, significant backlash.

% \subsection{The Baltic Incident}
% The first of these shootdowns, on April 1950, brought the previously-hidden aerial reconnaissance program to the attention of the American people. An unarmed Navy Patrol plane, a PB4Y2 Privateer, was intercepted and downed by Soviet fighters over the Baltic sea. The Soviet ambassador to the US lodged a formal note of protest, incorrectly alleging that the US aircraft which had been shot down was a B-29 bomber.\footcite{kirk_ambassador_1950} Upon receiving the note, Secretary of State Dean Acheson remarked that Ambassador Vishinsky's manner was ``serious but not aggressive or antagonistic,'' and the Secretary ``recommended that publicity on our side should be avoided or, if unavoidable, minimized.''\footcite{kirk_ambassador_1950}

% A few days later, the American ambassador issued a formal response, in which he correctly identified the model of plane and claimed that at no point did it cross into any territorial waters. He blamed the Soviet Union directly, outlining the punishment that the Americans planned to impose:

% % \begin{quote}
% 	The United States Government demands that the the Soviet Government institute a prompt and thorough investigation of this matter in order that the facts set forth above my be confirmed to its satisfaction. The United States Government further demands that the most strict and categorical instructions be issued to the Soviet air force that there be no repetition, under whatever pretext, of incidents of this kind which are so clearly calculated to magnify the difficulties of maintaining peaceful and correct international relationships.

% 	The United States Government fully expects that, when its investigation is completed, the Soviet Government will express its regret for the unlawful and provocative behavior of its aviators, will see to it that those responsible for this action are promptly and severely punished and will, in accordance with established custom among peace-loving nations, pay appropriate indemnity for the unprovoked destruction of American lives and property.\footcite{the_associated_press_text_1950}
% % \end{quote}
% This was a weak response---it imposed no consequences, threatened no sanctions, and demanded only the most \emph{pro forma} of reparations. In essence, the note demanded an apology and a promise not to repeat. There was absolutely no chance that the parties responsible on the Soviet side would be held accountable---in the most serious of cases, some lower-level officers might take the fall, but even that was unlikely to happen here. If the United States had followed through on its demand for reparations---the only material repercussion mentioned---then one could have argued that this statement had some teeth. Instead, the demands were rejected in totality by the USSR.\footcite{salisbury_kremlin_1950} As far as I can tell, no reparations were ever made and no further consequences were imposed as a result.

% Both sides had reasons to make this a larger political issue. The Soviet Union had fired first on a US Navy aircraft---an aircraft which could not have been the aggressor because its only armament was a .45-cal pistol carried by one of the crewmen.\footcite[p.~7]{peterson_maybe_1993} Given that the plane was never found, all available evidence suggests that the flight likely did not violate the traditional 12-mile airspace boundary. Based on the last received position report, today the CIA claims that the plane was flying 20-25 miles off the coast of Latvia at the time it was shot down.\footcite[p.~7]{peterson_maybe_1993} At the time, some unnamed Navy and Air Force officers, speaking to reporters in response to the Soviet protest note, conceded that it was possible that a complete breakdown in the plane's navigation system might have caused it to veer off-course into Latvia, but given its last known position, it would have taken ``a navigational error of nearly 90 degrees to cause the craft accidentally to wander over the Baltic states.''\footcite[Technically, the  article credits the first statement about the possible electronic breakdown to ``Navy and Air Force officials'' and the 90 degrees statement to ``Aerial navigators.'' It is unclear whether the navigators in questions are with the US military, but it does not matter regardless.]{the_new_york_times_soviet_1950}

% % \begin{figure}
% % \centering
% % \includegraphics[scale=0.45]{baltic-flight-path.png}
% % \caption{USN PB4Y2 Flight Path}
% % \label{baltic-flight-path}
% % \end{figure}

% Meanwhile, the United States knew that it hadn't lost a B-29, and that the missing plane probably hadn't ended up in Soviet airspace, but the Privateer in question \emph{had} been performing a questionably legal reconnaissance mission when it was shot down. Though the US was likely correct that the plane never flew over Latvia, it certainly got as close as it needed to in order to gather the necessary intelligence about Soviet military installations on the coast. No traditional boundaries were violated, but a repurposed Navy privateer loaded with electronic reconnaissance equipment flying close to Latvia looked, to use a less technical term, shady. Yet based on the initial exchange of diplomatic protests, neither side was inclined to magnify the issue if they could avoid it.

% The American press had no such reservations. \emph{Washington Post} columnist Walter Lippmann wrote that it could not have been ``a local incident brought on by a local commander but [rather] that it was an act of Soviet policy. The known facts indicate that Soviet intelligence \textelp{} believed it was carrying important electronic equipment and that orders were given to the Soviet fighter command to intercept it.''\footcite{lippmann_baltic_1950} He speculated that there was no way the plane could have violated Soviet airspace, because wreckage would have been recovered, the plane was too slow, and no commander would have sent it on such a dangerous mission intentionally. ``If, when, and as the American command were reconnoitering the Soviet military establishments on the Baltic coast,'' Lippmann wrote, ``it would use a plane of a wholly different type.''\footcite{lippmann_baltic_1950}

% A newspaper columnist is not a policymaker, but his argument was likely reflective of the political opinion that Americans held regarding espionage at the time: of course we do these things, we just do them \emph{better} than that---and if the US had been doing the thing that the Soviets claimed it was, then it would have been justified of them to open fire. ``Baltic Plane Mystery,'' a \emph{Post} article written a few days later, was almost sympathetic to the Soviet commanders. ``Electronics make the old delimitations for border coastal flights ridiculous. A plane flying on a course perfectly legal by standards accepted today might still be engaged in reconnaissance of the first importance that an unfriendly power would try to frustrate.''\footcite{childs_baltic_1950} This is true, both in terms of the mission itself and the technological environment, but he went even further. ``The only sound and safe assumption is that the Russians have a thorough and far-reaching espionage system. And at the same time we must hope that our system, and particularly on the side of counterespionage, is effective.'' Even the press, though outraged at the loss of American life, apparently took it as a given that this kind of espionage is necessary---and expected from both sides.

% The most important consequence of the Baltic Incident was not diplomatic. As a direct result of the shootdown, President Truman, who until that point was largely uninformed about these missions, ordered a thirty-day halt to all reconnaissance flights. In that time, his office drafted an operating procedure that would balance the strategic need for reconnaissance with its potential diplomatic volatility.\footcite[p.~41]{farquhar_aerial_2015} The resulting memo from the Joint Chiefs of Staff created, with Truman's approval, the Special Electronic Airborne Search Project (SESP), which set clear delineations of responsibility between Navy and Air Force reconnaissance, and established executive responsibility for approving the missions.

% Farquhar cites this as a landmark moment in the history of aerial reconnaissance. ``No longer would ferret operations be conducted ad hoc by the military services;'' he writes, ``from 1950 onward, reconnaissance operations attracted Presidential attention and played a significant role in shaping U.S. foreign policy.''\footcite[p.~42]{farquhar_aerial_2015} The strict procedures that SESP established were crafted to avoid the risk of escalation, while still explicitly acknowledging the ``serious disadvantages accruing to the United States if the cessation of these operations were to be extended over an excessively long period.''\footcite{bradley_memorandum_1950} But the SESP procedures also contain a curious provision: ``Flights will not be made closer than twenty miles to the USSR or USSR- or satellite-controlled territory.''\footcite{bradley_memorandum_1950} This rule would not be followed---as is typical of espionage, the consequences of the activity never outweigh its strategic value.

% \subsection{Routine resolution}
% Between the shootdown of the Navy Privateer in 1950 and the famous U-2 shootdown in 1960, the United States lost nine more aircraft to Soviet air defenses. Without analyzing each one at the same level of detail, I can demonstrate how the general attitudes displayed in response to the Baltic Incident became normalized. In each of the incidents listed, the pattern is the same---each side released dueling statements of protest, and the situation continued more or less as before.

% Describe here are the next few peacetime reconnaissance shootdowns whose diplomatic consequences Farquhar analyzes. A B-29 that went missing on October 7, 1952 hit the front page of the \emph{New York Times} two days later. This mission is on Peterson's list as a ferret flight. Farquhar notes that the incident happened during the US election season and that the media thought the attack was intended to lower American prestige, but he does not mention any diplomatic consequences.\footcite[p.~43-44]{farquhar_aerial_2015} The official US response, however, is preserved on the State Department website. As with the Baltic Incident, the US simply demanded that the Soviets pay for the plane and return any survivors.\footcite{the_new_york_times_u.s._1952} Two more shootdowns in March 1953---one involving an American F–84 Thunderjet and the other an RAF Lincoln bomber---actually resulted in conciliatory responses from both sides, including a secret meeting to reduce aerial tensions ``of which little is known''.\footcite[p.~45]{farquhar_aerial_2015}

% In another incident on July 29, 1953, the United States protested the shootdown of a B-50 (reconnaissance flight) over the Sea of Japan. The Soviet Union responded by protesting the alleged American shootdown of a Soviet passenger plane. Despite press outrage, no further action was taken.\footcite[p.~47]{farquhar_aerial_2015} A September 4, 1954 shootdown of a Navy P2V Neptune actually led to a call by one US Senator to break off diplomatic ties with the Soviet Union: ``Just another note from our State Department to the Kremlin hierarchy will not impress these uncivilized rulers \textelp{} that this new attack upon an American plane confirms Communist arrogance and aggressiveness to a point where the breaking of diplomatic relations is justified.''\footcite{the_associated_press_ending_1954} Instead of a harsher punishment, Eisenhower brought the issue to the UN with the full knowledge that an unfavorable judgment would be vetoed by the Soviets. The move demonstrated Eisenhower's ability to resist the calls for harsher action while still placating the American public.\footcite[p.~47]{farquhar_aerial_2015}

% Without belaboring the point, suffice to say that Farquhar concludes that after this moment in 1954, the resolution of these incidents became less tense. ``International incidents posed by shoot downs of reconnaissance aircraft still acted as a barrier in the path of d\'etente; but, by late 1954, overriding strategic concerns dictated a move toward breaking the cycle of hostility.''\footcite[p.~49]{farquhar_aerial_2015} In terms of media coverage, and diplomatic significance, no shootdown that followed would require more significant consequences than the ones that had already happened.

% One incident in the period between 1950-1954 does not fit the broader pattern. On November 20, 1951, an American C-47 transport was forced down in Hungary. This incident is omitted from Peterson's list because, despite Soviet cries of ``spies and saboteurs,'' there is no evidence that it was a reconnaissance flight.\footnote{It also wasn't exactly shot down so much as forced to land.} Of all the cases that Farquhar examines, this is the only one that provoked a sustained diplomatic response from the United States---and it is the only case where the plane in question definitively carried no espionage equipment. While the ferret flights were perfectly calibrated to fit as much radio surveillance equipment as possible, the only evidence of espionage that the Soviets were able to produce from the C-47---a plane that they recovered intact---was a portable radio, two extra parachutes, and some packets of warm blankets.\footcite{the_united_press_soviet_1951}
